---
title: "Quantile Binning for Microbiome ML"
subtitle: "EuroBioC2026 — Results Report"
author:
  - name: "Rasmus Hindström"
    affiliation: "Department of Computing, University of Turku"
    email: "rasmus.hindstrom@utu.fi"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    theme: flatly
    embed-resources: true   # single self-contained HTML
    code-fold: true
    code-tools: true
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
execute:
  echo: false
  warning: false
  message: false
  cache: false              # targets handles caching — never cache in Quarto
params:
  store: "../_targets"      # path to targets store, relative to this .qmd
---

```{r setup}
library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggsci)

tar_load(model_results)

# Order transformations so binning sits next to the others for easy comparison
trans_levels <- c("relabundance", "rclr", "pa", "binning")
results <- model_results |>
  mutate(
    transformation = factor(transformation, levels = trans_levels),
    is_binning     = transformation == "binning"
  )
```

# At-a-glance: does binning help?

The plots below are designed to answer one question: **does the binning
transformation deliver competitive test-set AUC compared to the standard
transformations (relabundance, rCLR, presence/absence)?** Each panel slices
the same `model_results` table from a different angle.

## Test AUC by transformation, per classifier

Each point is one dataset. Boxplots summarise the spread across datasets.
If binning works, its box should sit at or above the others.

```{r fig-box-auc, fig.height=5}
ggplot(results, aes(x = transformation, y = Accuracy, fill = is_binning)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.15, height = 0, size = 1.5, alpha = 0.7) +
  facet_wrap(~ method) +
  scale_fill_manual(values = c(`FALSE` = "grey70", `TRUE` = "#D7263D"),
                    guide = "none") +
  labs(x = NULL, y = "Accuracy",
       title = "Test Accuracy across datasets, by transformation and classifier") +
  theme_minimal(base_size = 12)
```

## Per-dataset comparison: binning vs. best alternative

For each dataset × classifier, we compute the gap between binning's AUC and
the best non-binning transformation. Positive bars mean binning *wins* on
that dataset; negative bars mean a standard transformation does better.

```{r fig-delta-auc, fig.height=6}
delta <- results |>
  select(dataset, method, transformation, Accuracy) |>
  pivot_wider(names_from = transformation, values_from = Accuracy) |>
  mutate(
    best_other = pmax(relabundance, rclr, pa, na.rm = TRUE),
    delta_auc  = binning - best_other
  )

ggplot(delta, aes(x = reorder(dataset, delta_auc), y = delta_auc,
                  fill = delta_auc > 0)) +
  geom_col() +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  coord_flip() +
  facet_wrap(~ method) +
  scale_fill_manual(values = c(`FALSE` = "grey60", `TRUE` = "#2E86AB"),
                    guide = "none") +
  labs(x = NULL, y = "Accuracy(binning) − Accuracy(best other)",
       title = "Where does binning beat the best alternative?") +
  theme_minimal(base_size = 12)
```

## Head-to-head scatter

Binning vs. each alternative, point-by-point. Points above the diagonal
mean binning wins on that (dataset, classifier) pair.

```{r fig-scatter, fig.height=5}
scatter <- results |>
  select(dataset, method, transformation, Accuracy) |>
  pivot_wider(names_from = transformation, values_from = Accuracy) |>
  pivot_longer(c(relabundance, rclr, pa),
               names_to = "alternative", values_to = "alt_auc")

ggplot(scatter, aes(x = alt_auc, y = binning, colour = method)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey50") +
  geom_point(size = 2, alpha = 0.8) +
  facet_wrap(~ alternative) +
  scale_color_jco() +
  coord_equal() +
  labs(x = "CV Accuracy (alternative transformation)",
       y = "CV Accuracy (binning)",
       title = "Binning vs. alternatives — points above the line favour binning") +
  theme_minimal(base_size = 12)
```

## Selected `n_bins` distribution

Which bin counts did the inner CV pick? A tight distribution suggests a
sweet spot; a flat one suggests `n_bins` doesn't matter much.

```{r fig-nbins, fig.height=4}
results |>
  filter(transformation == "binning") |>
  ggplot(aes(x = factor(selected_n_bins), fill = method)) +
  geom_bar(position = "dodge") +
  scale_fill_jco() +
  labs(x = "Selected n_bins (inner-CV winner)",
       y = "Number of datasets",
       title = "Tuned n_bins across datasets") +
  theme_minimal(base_size = 12)
```

## Summary table

```{r tbl-summary}
results |>
  group_by(method, transformation) |>
  summarise(
    n_datasets = n(),
    mean_auc   = mean(Accuracy, na.rm = TRUE),
    median_auc = median(Accuracy, na.rm = TRUE),
    sd_auc     = sd(Accuracy, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  arrange(method, desc(mean_auc)) |>
  knitr::kable(digits = 3)
```

