---
title: "Quantile Binning for Microbiome ML"
subtitle: "EuroBioC2026 — Results Report"
author:
  - name: "Rasmus Hindström"
    affiliation: "Department of Computing, University of Turku"
    email: "rasmus.hindstrom@utu.fi"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    theme: flatly
    embed-resources: true   # single self-contained HTML
    code-fold: true
    code-tools: true
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
execute:
  echo: false
  warning: false
  message: false
  cache: false              # targets handles caching — never cache in Quarto
params:
  store: "../_targets"      # path to targets store, relative to this .qmd
---

```{r setup}
library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggsci)

tar_load(model_results)

# Order transformations so binning sits next to the others for easy comparison
trans_levels <- c("relabundance", "rclr", "pa", "binning")
metric_cols  <- c("Accuracy", "Kappa")

results <- model_results |>
  mutate(
    transformation = factor(transformation, levels = trans_levels),
    is_binning     = transformation == "binning"
  )

results_long <- results |>
  select(dataset, method, transformation, is_binning, all_of(metric_cols)) |>
  pivot_longer(all_of(metric_cols), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric, levels = metric_cols))
```

# At-a-glance: does binning help?

The plots below are designed to answer one question: **does the binning
transformation deliver competitive test-set AUC compared to the standard
transformations (relabundance, rCLR, presence/absence)?** Each panel slices
the same `model_results` table from a different angle.

## Classification metrics by transformation, per classifier

Each point is one dataset. Boxplots summarise the spread across datasets.
If binning works, its box should sit at or above the others on every metric.

```{r fig-box-metrics, fig.height=8}
ggplot(results_long,
       aes(x = transformation, y = value, fill = is_binning)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.15, height = 0, size = 1.2, alpha = 0.7) +
  facet_grid(metric ~ method, scales = "free_y") +
  scale_fill_manual(values = c(`FALSE` = "grey70", `TRUE` = "#D7263D"),
                    guide = "none") +
  labs(x = NULL, y = "Metric value",
       title = "Accuracy and Kappa across datasets") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Per-dataset comparison: binning vs. best alternative

For each dataset × classifier, we compute the gap between binning's AUC and
the best non-binning transformation. Positive bars mean binning *wins* on
that dataset; negative bars mean a standard transformation does better.

```{r fig-delta-auc, fig.height=6}
delta <- results |>
  select(dataset, method, transformation, Accuracy) |>
  pivot_wider(names_from = transformation, values_from = Accuracy) |>
  mutate(
    best_other = pmax(relabundance, rclr, pa, na.rm = TRUE),
    delta_auc  = binning - best_other
  )

ggplot(delta, aes(x = reorder(dataset, delta_auc), y = delta_auc,
                  fill = delta_auc > 0)) +
  geom_col() +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  coord_flip() +
  facet_wrap(~ method) +
  scale_fill_manual(values = c(`FALSE` = "grey60", `TRUE` = "#2E86AB"),
                    guide = "none") +
  labs(x = NULL, y = "Accuracy(binning) − Accuracy(best other)",
       title = "Where does binning beat the best alternative?") +
  theme_minimal(base_size = 12)
```

## Head-to-head scatter

Binning vs. each alternative, point-by-point. Points above the diagonal
mean binning wins on that (dataset, classifier) pair.

```{r fig-scatter, fig.height=5}
scatter <- results |>
  select(dataset, method, transformation, Accuracy) |>
  pivot_wider(names_from = transformation, values_from = Accuracy) |>
  pivot_longer(c(relabundance, rclr, pa),
               names_to = "alternative", values_to = "alt_auc")

ggplot(scatter, aes(x = alt_auc, y = binning, colour = method)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey50") +
  geom_point(size = 2, alpha = 0.8) +
  facet_wrap(~ alternative) +
  scale_color_jco() +
  coord_equal() +
  labs(x = "CV Accuracy (alternative transformation)",
       y = "CV Accuracy (binning)",
       title = "Binning vs. alternatives — points above the line favour binning") +
  theme_minimal(base_size = 12)
```

## Selected `n_bins` distribution

Which bin counts did the inner CV pick? A tight distribution suggests a
sweet spot; a flat one suggests `n_bins` doesn't matter much.

```{r fig-nbins, fig.height=4}
results |>
  filter(transformation == "binning") |>
  ggplot(aes(x = factor(selected_n_bins), fill = method)) +
  geom_bar(position = "dodge") +
  scale_fill_jco() +
  labs(x = "Selected n_bins (inner-CV winner)",
       y = "Number of datasets",
       title = "Tuned n_bins across datasets") +
  theme_minimal(base_size = 12)
```

## Summary table

```{r tbl-summary}
results_long |>
  group_by(method, transformation, metric) |>
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = metric, values_from = mean_value) |>
  arrange(method, desc(Accuracy)) |>
  knitr::kable(digits = 3)
```

