---
title: "Quantile Binning for Microbiome ML"
subtitle: "EuroBioC2026 — Results Report"
author:
  - name: "Rasmus Hindström"
    affiliation: "Department of Computing, University of Turku"
    email: "rasmus.hindstrom@utu.fi"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    theme: flatly
    embed-resources: true   # single self-contained HTML
    code-fold: true
    code-tools: true
    fig-width: 8
    fig-height: 5
    fig-dpi: 150
execute:
  echo: false
  warning: false
  message: false
  cache: false              # targets handles caching — never cache in Quarto
params:
  store: "../_targets"      # path to targets store, relative to this .qmd
---

```{r setup}
library(targets)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggsci)

# Save a plot as a high-DPI PNG for the A0 poster and return it for rendering.
poster_fig <- function(plot, name,
                       width, height,
                       dpi = 300,
                       dir = here::here("analysis", "poster_figures")) {
  dir.create(dir, showWarnings = FALSE, recursive = TRUE)
  ggplot2::ggsave(
    filename = file.path(dir, paste0(name, ".png")),
    plot     = plot,
    width    = width,
    height   = height,
    dpi      = dpi,
    units    = "in"
  )
  plot
}

tar_load(model_results)
tar_load(tse_list_transformed)
tar_load(min_class_n)
tar_load(n_bins_levels)
tar_load(n_bins_range)
tar_load(cv_folds)

dataset_meta <- purrr::imap_dfr(tse_list_transformed, function(tse, nm) {
  counts_mat <- SummarizedExperiment::assay(tse, "counts")
  cd         <- as.data.frame(SummarizedExperiment::colData(tse))
  tibble::tibble(
    dataset    = nm,
    n_samples  = ncol(counts_mat),
    n_features = nrow(counts_mat),
    sparsity   = mean(counts_mat == 0),
    seq_type   = unique(cd$seq_type)[[1]],
    disease    = if ("disease" %in% names(cd)) unique(cd$disease)[[1]] else NA_character_
  )
})

# AUC is the primary metric (it is what inner-CV tuning optimises).
# Accuracy and Kappa are secondary.
trans_levels <- c("TSS", "rclr", "pa", "zQB")
metric_cols  <- c("AUC", "Accuracy", "Kappa")

results <- model_results |>
  mutate(
    transformation = if_else(transformation == "relabundance", "TSS", transformation),
    transformation = if_else(transformation == "binning", "zQB", transformation),
    transformation = factor(transformation, levels = trans_levels),
    is_binning     = transformation == "zQB"
  )

results_long <- results |>
  select(dataset, method, transformation, is_binning, all_of(metric_cols)) |>
  pivot_longer(all_of(metric_cols), names_to = "metric", values_to = "value") |>
  mutate(metric = factor(metric, levels = metric_cols))

method_colors <- c(
  rf      = "#5FA2CE",
  glmnet  = "#C85200",
  svm_rbf = "#57606C"
)

trans_colors <- c(
  TSS  = "#7B848F",
  rclr = "#57606C",
  pa   = "#5FA2CE",
  zQB  = "#C85200"
)

heatmap_low  <- "#5FA2CE"
heatmap_high <- "#C85200"
```

# Dataset overview

```{r callout-params, results="asis"}
# Emit a callout block with the pipeline parameters.
glue::glue("
::: {{.callout-note title='Pipeline parameters'}}
| Parameter | Value |
|---|---|
| Min samples per class (`min_class_n`) | {min_class_n} |
| n_bins candidates (`n_bins_levels`) | {n_bins_levels} (log₂-spaced) |
| n_bins range (`n_bins_range`) | {n_bins_range[1]}–{n_bins_range[2]} |
| CV folds (inner & outer) | {cv_folds} |
| Classifiers | Random Forest (`ranger`), Regularised logistic regression (`glmnet`), Radial SVM (`kernlab`) |
| Transformations | TSS, rCLR, presence/absence, quantile binning |
| Tuning metric | ROC-AUC (inner CV) |
:::
")
```

```{r tbl-datasets}
dataset_meta |>
  count(seq_type, name = "n_datasets") |>
  arrange(desc(n_datasets)) |>
  knitr::kable(
    col.names = c("Sequencing type", "Datasets"),
    caption   = paste0(
      "Datasets after filtering (min ", min_class_n,
      " samples per class): n = ", nrow(dataset_meta)
    )
  )
```

# Preprocessing pipeline

All datasets are agglomerated to genus level before any modelling. The table
below summarises which recipe steps are applied to each classifier, followed by
the rationale for each design decision.

```{r tbl-preprocessing}
tibble::tribble(
  ~Step,                         ~RF,  ~`glmnet`,  ~`svm_rbf`,
  "Genus-level agglomeration",   "✓",  "✓",        "✓",
  "Feature transformation",      "✓",  "✓",        "✓",
  "Zero-variance filter",        "✓",  "✓",        "✓",
  "Z-score normalisation",       "—",  "✓",        "✓"
) |>
  knitr::kable(caption = "Recipe steps by classifier class")
```

## Genus-level agglomeration

Raw count tables contain taxa resolved to varying depths. Agglomerating to
genus level reduces dimensionality, smooths over uncertain species-level
assignments, and gives each transformation a consistent feature space across
datasets.

## Feature transformations

Four transformations of the genus-level count matrix are compared:

- **Relative abundance (TSS)** — each sample divided by its library size.
  Corrects for sequencing depth; produces compositional proportions in $[0,1]$.
- **Robust centred log-ratio (rCLR)** — log-ratio of each count to the
  geometric mean of non-zero counts in that sample (zeros are handled without a
  pseudocount). Accounts for compositionality while preserving the zero structure.
- **Presence/absence (PA)** — binary indicator of detection. Discards all
  abundance information; useful when count magnitudes are unreliable or when
  only occupancy patterns matter.
- **Quantile binning** — non-zero counts within each sample are split into
  $n_\text{bins}$ equal-frequency quantile groups; zeros map to bin 0. The
  number of bins is a tunable hyperparameter selected by inner-CV AUC (see
  Section\ \@ref(binning-internals)). Because breaks are computed sample-wise,
  the transformation is leakage-free and can be applied inside the recipe.

All transformations are sample-wise: no cross-sample statistics are computed,
so they are safe to apply before the train/test split and inside inner CV folds
without leakage.

## Zero-variance filtering

`step_zv()` drops any feature whose values are identical across all training
samples in a given fold. This can occur after binning (many taxa collapse to
all-zero bins) and occasionally with presence/absence on extremely prevalent or
extremely rare taxa. Using exact zero-variance rather than the near-zero-variance
heuristic (`step_nzv`) avoids silently discarding rare-but-present taxa whose
low frequency ratios could still carry signal—a meaningful distinction in
sparse microbiome data.

## Z-score normalisation

`step_normalize()` centres and scales each feature to mean 0, variance 1 on the
training data of each fold. It is applied only to the three distance- or
penalty-sensitive classifiers:

**glmnet (elastic net):** The $L_1$ and $L_2$ penalties shrink coefficients
toward zero. Without standardisation, features on a larger numerical scale
receive a proportionally smaller effective penalty and are therefore
systematically favoured, which confounds the regularisation path. Standardisation
ensures that the penalty acts uniformly across features.

**svm_rbf (radial SVM):** The RBF kernel $k(\mathbf{x}, \mathbf{x}') =
\exp(-\sigma \|\mathbf{x} - \mathbf{x}'\|^2)$ measures Euclidean distance in
feature space. Features with large variances dominate the distance calculation
regardless of their biological relevance. Standardisation makes the kernel
sensitive to the relative, not absolute, spread of each feature.

::: {.callout-note title="Scaling ordinal bin labels"}
When the binning transformation is combined with z-score normalisation, the
recipe standardises ordinal integers (0, 1, 2, …, $n_\text{bins}$) rather than
continuous abundances. Two properties of the binned representation make this
well-behaved. First, the bins are **equally spaced** on an ordinal scale — each
step from bin $k$ to $k+1$ represents the same rank increase — so treating the
integers as a continuous scale is a reasonable approximation, and linear
centering/scaling preserves that spacing. Second, zeros (absent taxa) remain
distinguishable from bin 1 (lowest-abundance detected taxa) after scaling,
because 0 is a structurally distinct value that maps to a standardised score
well below the non-zero bin means. The practical consequence is that the
penalty (glmnet) or distance kernel (svm_rbf) sees a feature space where
each taxon's bin label has been placed on a common variance-1 scale — the same
regularisation assumptions that justify normalising continuous features apply
here.
:::

**Random Forest** does not receive normalisation. Decision trees split on
individual feature thresholds; they are invariant to any monotone transformation
of a feature's values, including centering and scaling. Adding normalisation to
RF would be harmless but adds unnecessary computation.

# Overall performance

Three classifiers are compared across all transformation × dataset combinations.
**Random Forest** (RF, `ranger`) is a bagging ensemble of decision trees; it handles
high-dimensional, sparse count data well and requires minimal preprocessing.
**Regularised logistic regression** (`glmnet`) fits an elastic-net penalised linear
model; features are standardised before fitting.
**Radial basis function SVM** (`kernlab`) uses an RBF kernel to separate classes in a
high-dimensional feature space; it also requires standardised inputs.

::: {.callout-note title="ROC-AUC" collapse="true"}
Area under the receiver operating characteristic curve. Measures the probability
that the model ranks a random case higher than a random control. Ranges from
0.5 (random) to 1.0 (perfect). **Used as the inner-CV tuning metric** because it
is threshold-independent and robust to class imbalance.
:::

::: {.callout-note title="Accuracy" collapse="true"}
Proportion of samples correctly classified. Intuitive, but sensitive to class
imbalance — a model that always predicts the majority class can score high without
being useful.
:::

::: {.callout-note title="Cohen's κ (Kappa)" collapse="true"}
Adjusts accuracy for the agreement expected by chance given the observed class
frequencies. Values near 0 indicate chance-level performance; values near 1 indicate
near-perfect discrimination. More informative than raw accuracy when classes are
imbalanced.
:::

Each point below is one dataset. Boxplots summarise the spread across all datasets.
AUC is shown first; Accuracy and Kappa follow in separate plots. Binning is
highlighted; its box should sit at or above the alternatives if it is competitive.

```{r fig-box-auc}
box_theme <- function() {
  list(
    theme_minimal(base_size = 20),
    theme(axis.text.x = element_text(angle = 0, hjust = 1),
          plot.title   = element_text(face = "bold"))
  )
}

poster_fig(
  results_long |>
    filter(metric == "AUC") |>
    ggplot(aes(x = method, y = value, fill = transformation)) +
    geom_boxplot(
      outlier.shape = NA, alpha = 1, colour = "black",
      position = position_dodge(width = 0.8)
    ) +
    scale_fill_manual(values = trans_colors, name = "Transformation") +
    labs(x = NULL, y = "AUC", title = "AUC by model x transformation") +
    box_theme(),
  name = "fig-box-auc", width = 10, height = 5
)
```

```{r fig-box-accuracy}
poster_fig(
  results_long |>
    filter(metric == "Accuracy") |>
    ggplot(aes(x = method, y = value, fill = transformation)) +
    geom_boxplot(
      outlier.shape = NA, alpha = 1, colour = "black",
      position = position_dodge(width = 0.8)
    ) +
    scale_fill_manual(values = trans_colors, name = "Transformation") +
    labs(x = NULL, y = "Accuracy", title = "Accuracy across datasets") +
    box_theme(),
  name = "fig-box-accuracy", width = 10, height = 5
)
```

```{r fig-box-kappa}
poster_fig(
  results_long |>
    filter(metric == "Kappa") |>
    ggplot(aes(x = method, y = value, fill = transformation)) +
    geom_boxplot(
      outlier.shape = NA, alpha = 1, colour = "black",
      position = position_dodge(width = 0.8)
    ) +
    scale_fill_manual(values = trans_colors, name = "Transformation") +
    labs(x = NULL, y = "Cohen's κ", title = "Kappa across datasets") +
    box_theme(),
  name = "fig-box-kappa", width = 10, height = 5
)
```

# Binning vs. PA

## Per-dataset comparison: binning vs. PA

For each dataset × classifier, we compute **Δ AUC = AUC(zQB) − AUC(PA)**. Red cells mean zQB wins on that dataset; blue cells mean a
standard transformation does better.

```{r fig-delta-auc, fig.height=6}
delta <- results |>
  select(dataset, method, transformation, AUC) |>
  pivot_wider(names_from = transformation, values_from = AUC) |>
  mutate(
    best_other = pmax(pa, na.rm = TRUE),
    delta_auc  = zQB - best_other
  ) |>
  group_by(dataset) |>
  mutate(mean_delta = mean(delta_auc, na.rm = TRUE)) |>
  ungroup() |>
  mutate(dataset = forcats::fct_reorder(dataset, mean_delta))

poster_fig(
  ggplot(delta, aes(x = method, y = dataset, fill = delta_auc)) +
    geom_tile(colour = "white", linewidth = 0.6) +
    geom_text(aes(label = sprintf("%+.3f", delta_auc)), size = 5) +
    scale_fill_gradient2(low = heatmap_low, mid = "white", high = heatmap_high,
                         midpoint = 0, name = "Δ AUC") +
    labs(x = NULL, y = NULL,
         title = "Where does zQB beat PA?") +
    theme_minimal(base_size = 20) +
    theme(panel.grid  = element_blank(),
          plot.title  = element_text(face = "bold")),
  name = "fig-delta-auc", width = 10, height = 12
)
```

## Head-to-head scatter

Binning AUC vs. each alternative, point-by-point. Points above the diagonal
mean binning wins on that (dataset, classifier) pair.

```{r fig-scatter, fig.height=5}
scatter <- results |>
  select(dataset, method, transformation, AUC) |>
  pivot_wider(names_from = transformation, values_from = AUC) |>
  pivot_longer(c(TSS, rclr, pa),
               names_to = "alternative", values_to = "alt_auc")

poster_fig(
  ggplot(scatter, aes(x = alt_auc, y = zQB)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "grey50") +
    geom_point(aes(colour = zQB > alt_auc), size = 2, alpha = 0.8) +
    facet_grid(method ~ alternative) +
    coord_equal() +
    scale_color_manual(values = c("TRUE" = "#457B9D", "FALSE" = "#BC4749"), guide = "none") +
    labs(x = "AUC (alternative transformation)",
         y = "AUC (zQB)",
         title = "zQB vs. alternatives — points above the line favour zQB") +
    theme_minimal(base_size = 20) +
    theme(plot.title = element_text(face = "bold")),
  name = "fig-scatter", width = 12, height = 14
)
```

# Binning internals

## Selected `n_bins` distribution

Which bin counts did the inner CV pick? A tight distribution suggests a
sweet spot; a flat one suggests `n_bins` doesn't matter much.

```{r fig-nbins, fig.height=7}
poster_fig(
  results |>
    filter(transformation == "zQB") |>
    ggplot(aes(x = factor(selected_n_bins), fill = method, colour = method)) +
    geom_bar(alpha = 1, position = position_dodge2(width = 0.9, preserve = "single")) +
    scale_fill_manual(values = method_colors, name = "Model") +
    scale_colour_manual(values = method_colors, guide = "none") +
    labs(x = "Selected n_bins (inner-CV winner)",
         y = "Number of datasets",
         title = "Tuned n_bins across datasets") +
    theme_minimal(base_size = 20) +
    theme(plot.title = element_text(face = "bold")),
  name = "fig-nbins", width = 10, height = 7
)
```


# Statistical comparison

Friedman test (non-parametric repeated-measures) treats each dataset as a
block and tests whether any transformation ranks consistently higher by AUC.
Pairwise Wilcoxon signed-rank tests (Holm-adjusted) follow up on the omnibus
result. Each cell shows the significance level and the mean AUC difference
(row − column); red cells favour the row transformation, blue cells favour
the column transformation.

```{r stat-tests}
trans_order <- c("relabundance", "rclr", "pa", "binning")
pairs       <- combn(trans_order, 2, simplify = FALSE)

test_wide <- model_results |>
  select(dataset, method, transformation, AUC) |>
  pivot_wider(names_from = transformation, values_from = AUC)

# Friedman test per classifier
friedman_tbl <- model_results |>
  select(dataset, method, transformation, AUC) |>
  group_by(method) |>
  group_modify(~ {
    ft <- friedman.test(AUC ~ transformation | dataset, data = .x)
    tibble::tibble(chi_sq  = round(ft$statistic, 2),
                   df      = ft$parameter,
                   p_value = ft$p.value)
  }) |>
  ungroup()

# Pairwise Wilcoxon (Holm-adjusted within each classifier)
pw_results <- test_wide |>
  group_by(method) |>
  group_modify(~ {
    purrr::map_dfr(pairs, function(p) {
      wt <- wilcox.test(.x[[p[1]]], .x[[p[2]]], paired = TRUE, exact = FALSE)
      tibble::tibble(trans1    = p[1],
                     trans2    = p[2],
                     mean_diff = mean(.x[[p[2]]] - .x[[p[1]]], na.rm = TRUE),
                     p_raw     = wt$p.value)
    }) |>
      mutate(p_adj = p.adjust(p_raw, method = "holm"))
  }) |>
  ungroup()
```

```{r tbl-friedman}
friedman_tbl |>
  knitr::kable(digits = c(0, 2, 0, 4),
               col.names = c("Classifier", "χ²", "df", "p-value"),
               caption   = "Friedman test: are transformation AUC ranks non-random?")
```

```{r fig-pw-wilcox, fig.height=5}
pw_sym <- pw_results |>
  bind_rows(
    rename(pw_results, trans1 = trans2, trans2 = trans1) |>
      mutate(mean_diff = -mean_diff)
  ) |>
  mutate(
    sig    = cut(p_adj,
                 breaks = c(0, 0.001, 0.01, 0.05, 1),
                 labels = c("***", "**", "*", "ns"),
                 right  = TRUE),
    trans1 = factor(trans1, levels = trans_order) |> forcats::fct_recode("zQB" = "binning", "TSS" = "relabundance"),
    trans2 = factor(trans2, levels = rev(trans_order)) |> forcats::fct_recode("zQB" = "binning", "TSS" = "relabundance"),
    label  = paste0(as.character(sig), "\n", sprintf("%+.3f", mean_diff))
  ) |>
  filter(
    match(as.character(trans1), c("TSS", "rclr", "pa", "zQB")) <
    match(as.character(trans2), c("TSS", "rclr", "pa", "zQB"))
  ) |>
  droplevels()

poster_fig(
  ggplot(pw_sym, aes(x = trans1, y = trans2, fill = mean_diff)) +
    geom_tile(colour = "white", linewidth = 0.6) +
    geom_text(aes(label = label), size = 5, lineheight = 0.9) +
    facet_wrap(~ method, ncol = 2) +
    scale_fill_gradient2(low  = heatmap_low,
                         mid  = "white",
                         high = heatmap_high,
                         midpoint = 0,
                         name = "Mean Δ AUC\n(row − col)") +
    labs(x = NULL, y = NULL,
         title   = "Pairwise Wilcoxon signed-rank tests (Holm-adjusted)") +
    theme_minimal(base_size = 20) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1),
          panel.grid  = element_blank(),
          plot.title  = element_text(face = "bold")),
  name = "fig-pw-wilcox", width = 14, height = 12
)
```

# Summary table

Mean [95% CI] for AUC, Accuracy, and Kappa per classifier × transformation across
all datasets, sorted by mean AUC descending. CIs are t-based across datasets.

```{r tbl-summary}
ci_fmt <- function(x) {
  n  <- sum(!is.na(x))
  mu <- mean(x, na.rm = TRUE)
  se <- sd(x, na.rm = TRUE) / sqrt(n)
  lo <- mu - qt(0.975, max(n - 1, 1)) * se
  hi <- mu + qt(0.975, max(n - 1, 1)) * se
  sprintf("%.3f [%.3f–%.3f]", mu, lo, hi)
}

auc_means <- results_long |>
  filter(metric == "AUC") |>
  group_by(method, transformation) |>
  summarise(mean_auc = mean(value, na.rm = TRUE), .groups = "drop")

results_long |>
  group_by(method, transformation, metric) |>
  summarise(fmt = ci_fmt(value), .groups = "drop") |>
  pivot_wider(names_from = metric, values_from = fmt) |>
  left_join(auc_means, by = c("method", "transformation")) |>
  arrange(method, desc(mean_auc)) |>
  select(-mean_auc) |>
  knitr::kable(
    col.names = c("Classifier", "Transformation", "AUC", "Accuracy", "Kappa"),
    caption   = "Mean [95% CI] across datasets (t-distribution, per classifier × transformation)"
  )
```
